Personalised Map Matched Imputation: imputing missing data from smartphone location logs

### 2018-05-08 Thesis Manuscript/manuscript.tex
final commit

    Submitted as a research report conforming to APA manuscript guidelines
    (6th edition).

@@

  \abstract{Personal mobility, or how people move in their environment, is
associated with a vast range of behavioural traits and outcomes, such as
socioeconomic status, personality and mental health. The widespread
adoption of location-sensor equipped smartphones has generated a wealth
of objective personal mobility data. Nonetheless, smartphone collected
personal mobility data has remained underutilised in behavioural
research, partly due to the practical difficulties associated with
obtaining the data and partly because of the methodological complexity
associated with analysing it. Recent changes in European regulation have
made it easier for researchers to obtain this data, but the
methodological difficulties remain. The difficulty lies in that
smartphone location data is irregularly sampled, sparse and often
inaccurate. This results in a high proportion of missing data and
significant noise. In this paper we present a method called Personal Map
Matched Imputation (PPMI) to deal with missing data and noise in
smartphone location logs. The main innovation of PPMI is that it creates
a personalised spatial map for each individual based on all the
available data. In doing so PPMI leverages the regularity of human
mobility in order to smoothen noisy measurments and impute missing data
values. By simulating missing periods in real data we find that a simple
implementation of PPMI performs as well as existing methods for short (5
minute) missing intervals and substantially better for longer (1 day)
missing intervals. When imputing a subset of real missing data where
travel logs are available as a reference points, we find that PPMI
performs substantially better than existing models.}
  \keywords{Missing Data, Measurement Bias, GPS, Human Mobility \\

@@

Why is human mobility important? Human mobility measures are quantified
metrics of how people move about in their environment. Human mobility
affects a wide range of outcomes, such as health, income and social
capital (Goodchild \& Janelle, 2010). The most widely administered
personality questionnaires ask individuals to what extent they agree
with statements such as \enquote{I love large parties}, \enquote{I
prefer going to the movies to watching videos at home} and \enquote{I
love to travel to places that I have never been before} (Goldberg et
al., 2006). At their root these questions are mobility measures. In
economic research, the postal code of an individual's home address is
often used as a proxy for socioeconomic status (e.g. Villanueva \&
Aggarwal, 2013). Economists are interested not only in where people
live, but also where they go to work: geographic labour mobility is
related to the income of individuals, the well-being of a country's
economy and even informs the policy of bodies such as the European
Comission (Tatsiramos, 2009). The extensive use of measures like these
across different domains in social science strongly suggests that
mobility metrics are linked to important real world outcomes. Perhaps it
is unsurprising that behavioural researchers have found that mobility
measures can be used to predict academic performance (R. Wang, Harari,
Hao, Zhou, \& Campbell, 2015), the incidence of obesity (Zenk, Schulz,
\& Odoms-Young, 2009) or even the onset of a depressive episode in
bipolar depression patients (Palmius et al., 2017). Indeed, there is an
argument to be made that perhaps psychologists have not been studying
mobility enough. When studying behavioural differences within
individuals behavioural scientists have often neglected the fact that
individuals vary not only over time but also over space. To fully
understand behaviour we must understand how behaviour can vary across
environments.

Despite the importance of mobility measures, the majority of these
metrics are obtained through the use of questionnaires (such as the
aforementioned personality questionnaire) or specifically through
pen-and-paper travel diaries. As a thought experiment we encourage the
reader to try to remember where exactly he or she was five weeks ago on
Saturday at midday. Most people find this sort of task very difficult
(if not impossible), which illustrates that questionnaires and travel
diaries are have well-known methodological flaws. This method of
collecting data is difficult and relies on accurate self-reporting.
Travel diaries are burdensome to collect because participants must be
explicitly asked to write down on their movement patterns at frequent
intervals due to human forgetfulness. This makes the data collection
expensive and limits the duration of data collection in practice. In
addition, the frequent reporting duties of the participant may bias the
participants behaviour. The limits of human cognition also limit the
accuracy of self-reported measures. There is evidence that participants
are systematically biased when self-reporting mobility measures. For
instance, participants under-report the frequency of short trips (Wolf,
Oliveira, \& Thompson, 2003) and underestimate the duration of regular
commutes (Delclòs-Alió, Marquet, \& Miralles-Guasch, 2017). These
obstacles can only be overcome by using objective data on human
mobility.

Social scientists now have the unprecedented opportunity to easily
obtain objective data on human mobility from smartphones. Before
smartphones the only way to collect objective data on human mobility
involved giving participants an expensive professional-grade location
sensor and convincing them to take it with themselves at all times.
Barnett and Onnela (2016) points out that introducing a new device to
the participant's life may bias their behaviour. Moreover, collecting
data in such a way is costly, places a high burden on participants and
therefore the logs do not span a long time (Barnett \& Onnela, 2016).
Today millions of individuals carry smartphones with themselves every
day and do not need to be encouraged to do so by researchers.
Smartphones are equipped with a range of sensors that can be used to
track the location of the device at all times. These smartphones can
collect and store hundreds of location measurements a day. For instance,
Google Location History contains movement information on millions of
users, often spanning years (Location History, 2017). Moreover, recent
changes in EU regulations with regard to consumer data-portability
rights ensure that a willing participant should be able to easily share
this information with researchers at no cost to the either the
participant or the researchers (Commission, 2017). Taken together, this
means that researchers now have at their disposal the ability to easily
access the objective human mobility data of millions of individuals
spanning several years at little cost and without a significant burden
of the participants.

@@

Smartphone location measures are obtained primarily (but not
exclusively) by Global Positioning System (GPS) measurements. A GPS
sensor uses the distance between a device and several satellites to
determine the location of the device. Although using a GPS sensor is the
most accurate way to establish location on a smartphone, the GPS sensor
is also the most energy consuming sensor on most smartphones (M. Y. Chen
et al., 2006; LaMarca et al., 2005). In order to avoid battery depletion
and to overcome computational constraints smartphones also use
less-accurate heuristics such as WiFi access points and cellphone tower
triangulation. Smartphone location logs contain measurements from all of
these sources, usually in the form of time-stamped latitude, longitude
and accuracy values. The accuracy \(a\) of any given measurement is
given in meters such that it represents the radius of a 67\% confidence
circle (Location History, 2017). In other words, the true location of a
device should be within the radius \(a\) of the measurement 67\% of the
time.

Researchers often develop custom-made tracking applications which
participants are instructed to download on their phone. Alternatively
participants are given a phone to use for a given period of time with
the custom-made tracking app pre-installed. We call location logs
resulting from these custom-made apps \emph{custom logs}. The advantage
of custom logs is that the researchers can adjust tracking parameters,
such as the the frequency of measurements and the sensor with which they
are made. The disadvantage with this approach is that researchers have
to develop or adapt a custom-made tracking application (which is not
easy given hundreds of different types of smartphone models), distribute
it among research participants and enforce participation. Participants
may dislike tracking apps because they view them as more intrusive and
these apps regularly drain the battery of the device (G. M. Harari et
al., 2016). Moreover, researchers have distribute this application among
research participants and convince them not to turn it off.

We focus on another solution, which is to take advantage of existing
smartphone location logs (\emph{secondary logs}) . The advantages are
clear: repositories such as Google's Location History contains
information on millions of users spanning years (Location History,
2017), participants can share the data by the click of a button and
there can be no behavioural changes due to participation in the study as
the participant share past data. The disadvantage is that researchers
have no control over the tracking parameters, often resulting in logs
with sparse and inaccurate measurements. Hence, two important challenges
are dealing with missing data and measurement noise.

In order to work with secondary logs, researchers need to be able to
handle the data sparsity that leads to missing data. Missing data is a
pervasive issue in secondary logs as it can arise due to several
reasons. Technical reasons include signal loss, battery failure and
device failure. Behavioural reasons include leaving the phone at home or
switching the device off. As a result, secondary logs often contain wide
temporal gaps with no measurements. For instance, several research
groups studying mental health report missing data rates between 30\% to
50\% (Grünerbl et al., 2015; Palmius et al., 2017; Saeb et al., 2015).
Other researchers report similar trends in different fields (e.g. G. M.
Harari et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015). In Figure
\ref{fig:longMeasurementsPerDay} shows that despite the long duration of
the log the sparsity it is also evident.

@@

There is no golden standard for dealing with missing data in GPS logs
(Barnett \& Onnela, 2016). Importantly, spatiotemporal data measurements
are auto-correlated in both time and space. This means that best
practices with other types of data, such as mean imputation, are
unsuitable. For example, imagine an individual who splits almost all her
time between work and home. Suppose she spends a small amount of time
commuting between the two along a circular path. Using mean imputation
to estimate her missing coordinates, we impute her to be at the midpoint
between home and work, even though she has never been there. Worryingly,
there is little transparency on how researchers deal with missing data
(Jankowska et al., 2015).

Another methodological problem is related to the noise in the
measurements that are collected. The accuracy of smartphone location
measurements is substantially lower than that of professional GPS
location trackers because smartphones often use less accurate sensors.
In professional GPS trackers less than 80\% of measurements fall within
10 meters of the true location. GPS measures are most inaccurate in
dense urban locations and indoors (S. Duncan et al., 2013; Schipperijn
et al., 2014). Unfortunately for researchers, this is where people in
the developed world spend most of their time. Figure
\ref{fig:accuracyPlot} shows how accuracy can vary as a function of user
behaviour, time and location. Most notably, low accuracy is often (but
not always) associated with movement (see Figure
\ref{fig:accuracyPlot2}).

@@

Noisy data can lead to inaccurate conclusions if it is not accounted
for. Suppose we wish to calculate an individual's movement in a day. A
simple approach would be to calculate the sum of the distance between
each measurement. But if there is noise, the coordinates will vary even
though the individual is not moving. If the measurements are frequent
and noisy, we will calculate a lot of movement, even if the individual
did not move at all! This issue is also visualised in Figure
\ref{fig:palmiusVsMe1}. The problem is further complicated because
missing data and noisy measurements are related. Methods used by
researchers to reduce noise, such as throwing out inaccurate
measurements (e.g. Palmius et al., 2017), can exacerbate the severity of
the missing data problem.

In this paper we will propose PPMI as a method for dealing with missing
data and measurement error in secondary location logs. We will compare
PPMI to similar solutions in the literature by evaluating the distance
between points which were simulated as missing and their imputed
counterparts. Finally we will calculate time spent at home as a function
of the imputation method.

@@

How have researchers dealt with missing data in human mobility logs thus
far? Unfortunately there is no golden standard in how to deal with this
type of missing data. Researchers are generally vague about what
practices they follow (Jankowska et al., 2015). This vagueness is
worrisome as it invites solutions which contain significant researcher
degrees of freedom (Simmons, Nelson, \& Simonsohn, 2011). The vagueness
is possibly also due to the fact that most researchers are unfamiliar
with possible solutions. Most researchers simply down-sample temporally
and remove missing observations or use some sort of rule-based common
sense imputations (e.g. Palmius et al. (2017)). The only principled
approach that we know of that aims to solve the issue of missing data in
location logs as they relate to human mobility is that of Barnett and
Onnela (2016). We will explore the methods of Barnett and Onnela (2016)
and Palmius et al. (2017) in detail subsequently, after introducing
exploring other spatiotemporal methods.

A lack in methods for missing data imputation for human mobility
patterns does not imply there is not a vast literature on modelling
movement. The most widespread models are SSMs, therefore we shall detail
a few examples and subsequently argue that they are nonetheless unsuited
for long term human mobility logs. Ecologists have used SSMs to explain
how animals interact with their environment (T. A. Patterson, Thomas,
Wilcox, Ovaskainen, \& Matthiopoulos, 2008). These models can be quite
complex. Preisler, Ager, Johnson, and Kie (2004) uses Markovian movement
processes to characterise the effect of roads, food patches and streams
on cyclical elk movements. The most well studied SSM is the Kalman
filter, which is the optimal algorithm for inferring linear Gaussian
systems. The extended Kalman filter is the de facto standard for GPS
navigation (Z. Chen \& Brown, 2013). The advantage of state space models
is that they are flexible, deal with measurement inaccuracy, include
information from different sources and can be used in real time.

For secondary logs the main limitation of SSM implementations is that
they ignore movement routines. For instance, humans tend to go to work
on weekdays and sleep at night. Because SSMs are based on the Markov
property, they cannot incorporate this information. In other words, the
estimated location \(G(t)\) at time-point \(t\) is often based only upon
measurements \(D_t\), \(D_{t-1}\) and ignores all \(D_{t-i}|i\geq2\).
Hierarchical structuring and conditioning on a larger context have been
suggested as ways to add periodicity to Markovian models. These
solutions are often computationally intractable or unfeasible (Sadilek
\& Krumm, 2016). Moreover, these models often assume time and space
invariance (location is not a direct function of time or space). These
mathematical assumptions are violated in the case of human movement
patterns. For this reason we do not consider existing SSMs to be useful
for imputing missing data in this case.

@@

On the other hand, a few researchers have explicitly attempted to impute
missing data from human mobility patterns (Barnett \& Onnela, 2016;
Palmius et al., 2017 ; @ Wu, Fu, Wang, Xiao, \& Fu, 2017). Importantly,
none of them worked with secondary logs. Nonetheless we will detail what
they did as informative examples. Palmius et al. (2017) deal with the
measurement inaccuracy of \(D\) in custom logs by removing from the data
set all unique low-accuracy \(a\) data points that had
\(\frac{d}{dt}D > 100 \frac{km}{h}\). Subsequently the researchers down
sample the data to a sample rate of 12 per hour using a median filter.
Moreover, Palmius et al. (2017) explain:

\begin{quote}
\enquote{If the standard deviation of {[}\(D\){]} in both latitude and
longitude within a 1 h epoch was less than 0.01 km, then all samples
within the hour were set to the mean value of the recorded data,
otherwise a 5 min median filter window was applied to the recorded
latitude and longitude in the epoch}.
\end{quote}

Missing data was imputed using the mean of measurements close in time if
the participant was recorded within 500m of either end of a missing
section and the missing section had a length of \(\leq 2h\) or
\(\leq 12h\) after 9pm. In cases where the previous conditions are not
met no values are imputed.

Barnett and Onnela (2016) follow a different approach which is, to the
best of our knowledge, the only principled approach to dealing with
missing data in human mobility data. Barnett and Onnela (2016) work with
custom logs where location is measured for 2 minutes and subsequently
not measured for 10 minutes. In the words of the authors, Barnett and
Onnela (2016) handle missing data by first converting data to mobility
traces, which are defined as a sequence of flights and pauses. Flights
are segments of linear movements and pauses corresponding to periods of
time where a person does not move. Subsequently, the authors impute
missing data by:

@@

Subsequently let us simplify by assuming that a persons location is on
two-dimensional Euclidean plane. Let a person's true location on this
two-dimensional plane be \(G(t) = [G_x(t) G_y(t)]\) where \(G_x(t)\) and
\(G_y(t)\) denote the location of the individual at time \(t\) on the
x-axis and y-axis respectively. For simplicity, we can assume that the
x-axis is the longitude and the y-axis is the latitude. Moreover, let
\(D \in \mathbb{R}^2\) be the recorded data containing the latitude and
longitude. In addition, let \(a\) denote the estimated accuracy of the
recorded data. Furthermore, \(G(t)\), \(D\) and \(a\) are indexed by
time labeled by the countable set \(t = t_1 < ... < t_{n+1}\). The
measure of accuracy \(a_t\) is given in meters such that it represents
the radius of a 67\% confidence circle. If \(D_t = \emptyset\) it is
considered \emph{missing} and it is not missing otherwise.

When several data sets are available from individuals living in
overlapping areas we can construct a \(t \times i\) matrix \(M\) where
the entry \(M(t,i)\) contains \(G(t)\) for the individual \(i\).

\subsection{Personalised Map Matching
Imputation}\label{personalised-map-matching-imputation}

@@

First, following Barnett and Onnela (2016) we categorise all time-points
\(t\) as either belonging to the set \(P\) (pause) or set \(F\)
(flight). Conceptually pauses can be understood as periods of time where
an individual spends significant amount of continuous time without
moving. Flights are the times where the individual is moving. Let
\(t_a\) be a pause of length \(n\). \[t_a = t_i < ... < t_{i+n}\] Let
\(t_b\) be a pause of length \(m\) such that there is no temporal
overlap between \(t_a\) and \(t_b\):
\[t_b = t_j= < ... < t_{j+m}|t_{i+n}< t_j\] Then it follows that between
the two pauses there must be a flight indexed by \(t_x\) of length
\(j-i+n\). \[t_x =  t_{i+n} < ... < t_{j} |t_x \in F\] We define pause
locations \(G(t_a), G(t_b) | t_a, t_b \in P\) as locations where an
individual spends an extended amount of time in the same space
(e.g.~school, home, work, train station, barber shop, bar, gym).
Importantly, our model assumes period and cyclic human movement such
that there are many pauses \(t_{a1},t_{a2},...,t_{an}\) such that
\(G(t_{a1}) = G(t_{a2}) = ... = G(t_{an})\). Moreover,it is possible for
\(G(t_a) = G(t_b)\) such that \(t_a \ne t_b\). For example, if the
individual leaves home for a run and returns home without stopping
anywhere else.

@@

\subsection{Personalised Map Matching Imputation
algorithm}\label{personalised-map-matching-imputation-algorithm}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \emph{Map building}: Extract from measurements \(D\) all pause
  location bins and path location bins to create a personalised map.
\item
  \emph{Binning}: Assign each measurement \(D\) to a unique discrete
  location bin.
\item
  \emph{Imputing}: Use a classification method to predict missing
  measurements based on all the available information.
\end{enumerate}

@@

Following Wu et al. (2017)'s spatial-temporal-semantic (STS) feature
extraction algorithm our aim is to transform pause and path locations
into machine learning friendly discrete location sequences. There are
multiple ways of extracting such measurement clusters in the literature,
such as Spatio-Temporal Density-Based Spatial Clustering of Applications
with Noise (ST-DBSCAN) and sequence oriented clustering (SOC)
(``ST-DBSCAN,'' 2007). We will focus on two methods which explicitly
with mobility patterns from unevenly sampled smartphone logs (Barnett \&
Onnela, 2016; Palmius et al., 2017 ). Both of these methods pre-process
the data and subsequently use two steps to extract pause locations:
first they extract pauses and their corresponding locations, then they
cluster pause locations based on spatial proximity. This implementation
of PMMI uses a stricter version of Barnett and Onnela (2016)'s approach
to extract pauses.

@@

All the code is available on a public repository (Sobrado, 2018).

>>>>

    Submitted as a thesis manuscript conforming to APA manuscript guidelines
    (6th edition).

@@

  \abstract{Personal mobility, or how people move in their environment, is
associated with a vast range of behavioural traits and outcomes, such as
socioeconomic status, personality and health. The widespread adoption of
location-sensor equipped smartphones has generated a wealth of objective
personal mobility data. Nonetheless, smartphone collected personal
mobility data has remained underused in behavioural research, partly due
to the practical difficulties associated with obtaining the data and
partly because of the methodological complexity associated with
analysing it. Recent changes in European regulation have made it easier
for researchers to obtain this data, but the methodological difficulties
remain. The difficulty lies in that smartphone location data is
irregularly sampled, sparse and often inaccurate. This results in a high
proportion of missing data. In this paper we present a method called
Personal Map Matched Imputation (PPMI) to deal with missing data and
noise in smartphone location logs. The main innovation of PPMI is that
it creates a personalised spatial map for each individual based on all
the available data. In doing so PPMI leverages the regularity of human
mobility in order to smoothen noisy measurements and impute missing data
values. By simulating missing periods in real data we find that a simple
implementation of PPMI performs as well as existing methods for short (5
minute) missing intervals and substantially better for longer (1 day)
missing intervals. When imputing a subset of real missing data where
travel logs are available as a reference points, we find that PPMI
performs substantially better than existing models.}
  \keywords{Missing Data, Measurement Bias, GPS, Human Mobility \\

@@

Why is human mobility important? Human mobility affects a wide range of
outcomes, such as health, income and social capital (Goodchild \&
Janelle, 2010). Human mobility measures are quantified metrics of how
people move about in their environment. The most widely administered
personality questionnaires ask individuals to what extent they agree
with statements such as \enquote{I love large parties}, \enquote{I
prefer going to the movies to watching videos at home} and \enquote{I
love to travel to places that I have never been before} (Goldberg et
al., 2006). At their root these questions are mobility measures. In
economic research, the postal code of an individual's home address is
often used as a proxy for socioeconomic status (e.g. Villanueva \&
Aggarwal, 2013). Economists are interested not only in where people
live, but also where they go to work: geographic labour mobility is
related to the income of individuals, the well-being of a country's
economy and even informs the policy of bodies such as the European
Comission (Tatsiramos, 2009). The extensive use of measures like these
across different domains in social science strongly suggests that
mobility metrics are linked to important real world outcomes.
Behavioural researchers have found that mobility measures can be used to
predict academic performance (R. Wang, Harari, Hao, Zhou, \& Campbell,
2015), the incidence of obesity (Zenk, Schulz, \& Odoms-Young, 2009) or
even the onset of a depressive episode in bipolar depression patients
(Palmius et al., 2017). To fully understand behaviour not only must we
understand how behaviour can vary within and between individuals, but
also across environments.

Despite the importance of mobility measures, the majority of metrics are
obtained through the use of questionnaires (such as the aforementioned
personality questionnaire) or specifically through pen-and-paper travel
diaries. As a thought experiment we encourage the reader to try to
remember where exactly he or she was five weeks ago on Saturday at
midday. Most people find this sort of task difficult (if not
impossible), which illustrates the main methodological flaws of
questionnaires and travel diaries. Travel diaries are burdensome to
collect because participants must be explicitly asked to write down on
their movement patterns at frequent intervals (lest they forget them).
The burden to the participants makes data collection expensive, which in
turn limits the duration of data collection in practice. In addition,
the frequent reporting duties of the participant may bias the
participant's behaviour. The limits of human cognition also limit the
accuracy of self-reported measures. There is evidence that statistics
derived from self-reported mobility measures are systematically biased.
For instance, participants under-report the frequency of short trips
(Wolf, Oliveira, \& Thompson, 2003) and underestimate the duration of
regular commutes (Delclòs-Alió, Marquet, \& Miralles-Guasch, 2017).
These obstacles can only be overcome by using objective data on human
mobility.

Social scientists now have the unprecedented opportunity to easily
obtain objective data on human mobility from smartphones. Before
smartphones the only way to collect objective data on human mobility
involved giving participants an expensive professional-grade location
sensor and convincing them to take it with themselves at all times.
Barnett and Onnela (2016) points out that introducing a new device to
the participant's life may bias their behaviour. Moreover, collecting
data in such a way is costly, places a high burden on participants and
therefore the logs do not span a long time (Barnett \& Onnela, 2016).
Today millions of individuals carry smartphones with themselves every
day and do not need to be encouraged to do so by researchers.
Smartphones are equipped with a range of sensors that can be used to
track the location of the device at all times. These smartphones can
collect and store hundreds of location measurements a day. For instance,
Google Location History contains movement information on millions of
users, often spanning years (\emph{Timeline}, 2017). Moreover, recent
changes in EU regulations with regard to consumer data-portability
rights ensure that a willing participant should be able to easily share
this information with researchers at no cost to the either the
participant or the researchers (European Commission, 2017). Researchers
now have the ability to easily access the objective movement records of
millions of individuals spanning several years at little cost and
without a significant burden to the participants.

@@

Smartphone location measures are obtained primarily (but not
exclusively) by Global Positioning System (GPS) measurements. A GPS
sensor uses the distance between a device and several satellites to
determine the location of the device. Although using a GPS sensor is the
most accurate way to establish location on a smartphone, the GPS sensor
is also the most energy consuming sensor on most smartphones (M. Y. Chen
et al., 2006; LaMarca et al., 2005). In order to avoid battery depletion
and to overcome computational constraints smartphones also use
less-accurate heuristics such as WiFi access points and cellphone tower
triangulation. Smartphone location logs contain measurements from all of
these sources, usually in the form of time-stamped latitude, longitude
and accuracy values. The accuracy \(a\) of any given measurement is
given in meters such that it represents the radius of a 67\% confidence
circle (\emph{Timeline}, 2017). In other words, the true location of a
device should be within the radius \(a\) of the measurement 67\% of the
time.

Researchers often develop custom-made tracking applications which
participants are instructed to download on their phone. Alternatively
participants are given a phone to use for a given period of time with
the custom-made tracking app pre-installed. We call location logs
resulting from these custom-made apps \emph{custom logs}. The advantage
of custom logs is that the researchers can adjust tracking parameters,
such as the the frequency of measurements and the sensor used to make
the measurement. The disadvantage of this approach is that researchers
have to develop or adapt a custom-made tracking application (which is
not easy given hundreds of different types of smartphone models),
distribute it among research participants and enforce participation.
Participants may dislike tracking apps because they view them as more
intrusive and these apps regularly drain the battery of the device
(Harari et al., 2016).

We focus on another solution, which is to take advantage of existing
smartphone location logs (\emph{secondary logs}). The advantages are
clear: repositories such as Google's Location History contain years of
information on millions of users (\emph{Timeline}, 2017) and
participants can share the data by clicking of a button. The
disadvantage is that researchers have no control over the tracking
parameters, often resulting in logs with sparse and inaccurate
measurements. Hence, two important challenges in using secondary logs
are how to deal with missing data and measurement noise.

In order to work with secondary logs, researchers need to be able to
handle the data sparsity that leads to missing data. Missing data is a
pervasive issue in secondary logs as it can arise due to several
reasons. Technical reasons include signal loss, battery failure and
device failure. Behavioural reasons include switching the device or its
location services off. As a result, secondary logs often contain wide
temporal gaps with no measurements. For instance, several research
groups studying mental health report missing data rates between 30\% to
50\% (Grünerbl et al., 2015; Palmius et al., 2017; Saeb et al., 2015).
Other researchers report similar trends in different fields (e.g. Harari
et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015). Figure
\ref{fig:longMeasurementsPerDay} shows a secondary log spanning years
where despite its long duration the sparsity of measurements is also
evident.

@@

There is no golden standard for dealing with missing data in GPS logs
(Barnett \& Onnela, 2016). Importantly, spatiotemporal data measurements
are auto-correlated in both time and space. This means that common
practices with other types of data, such as mean imputation, are
unsuitable. For example, imagine an individual who splits almost all her
time between work and home. Suppose she spends a small amount of time
commuting between the two along a circular path. Using mean imputation
to estimate her missing coordinates, we impute her to be at the midpoint
between home and work, even though she has never been there. Worryingly,
there is little transparency on how researchers deal with missing data
(Jankowska et al., 2015).

Another methodological problem is related to the noise in the
measurements. The accuracy of smartphone location measurements is
substantially lower than that of professional GPS location trackers
because smartphones often use less accurate sensors. In professional GPS
trackers less than 80\% of measurements fall within 10 meters of the
true location. GPS measures are most inaccurate in dense urban locations
and indoors (S. Duncan et al., 2013; Schipperijn et al., 2014).
Unfortunately for researchers, this is where people in the developed
world spend most of their time. Figure \ref{fig:accuracyPlot} shows how
accuracy can vary as a function of user behaviour, time and location.
Most notably, low accuracy is often (but not always) associated with
movement (see Figure \ref{fig:accuracyPlot2}).

@@

Noisy data can lead to inaccurate conclusions if it is not accounted
for. Suppose we wish to calculate an individual's movement in a day. A
simple approach would be to calculate the sum of the distances between
each measurement. But if there is noise, the coordinates will vary even
though the individual is not moving. If the measurements are frequent
and noisy, we will calculate a lot of movement, even if the individual
did not move at all! The problem is further complicated because missing
data and noisy measurements are related. Methods used by researchers to
reduce noise, such as throwing out inaccurate measurements (e.g. Palmius
et al., 2017), can exacerbate the severity of the missing data problem.

In this paper we will propose PPMI as a method for dealing with missing
data and measurement error in secondary location logs. PPMI creates a
discrete personalised map based on the individuals mobility history.
Then it assigns each measurement to a discrete location based on the
individuals entire mobility history. Finally, it uses a classification
method to impute missing periods.

@@

How have researchers dealt with missing data in human mobility logs thus
far? Unfortunately there is no golden standard in how to deal with this
type of missing data. Researchers are generally vague about what
practices they follow (Jankowska et al., 2015). This vagueness is
worrisome as it invites solutions which contain significant researcher
degrees of freedom (Simmons, Nelson, \& Simonsohn, 2011). Most
researchers simply down-sample temporally and remove missing
observations or use some sort of rule-based common sense imputations
(e.g. Palmius et al., 2017). The only principled approach (that we know
of) aiming to solve the issue of missing data in this context
(i.e.~location logs as they relate to human mobility) is that of Barnett
and Onnela (2016). We will explore the methods of Barnett and Onnela
(2016) and Palmius et al. (2017) in detail subsequently, after
introducing other spatiotemporal methods.

A lack in methods for missing data imputation for human mobility
patterns does not imply there is not a vast literature on modelling
movement. The most widespread models are SSMs, therefore we shall detail
a few examples and subsequently argue that they are nonetheless unsuited
for long term human mobility logs. Ecologists have used SSMs to explain
how animals interact with their environment (T. A. Patterson, Thomas,
Wilcox, Ovaskainen, \& Matthiopoulos, 2008). Preisler, Ager, Johnson,
and Kie (2004) uses Markovian movement processes to characterise the
effect of roads, food patches and streams on cyclical elk movements. The
most well studied SSM is the Kalman filter, which is the optimal
algorithm for inferring linear Gaussian systems. The extended Kalman
filter is the de facto standard for GPS navigation (Z. Chen \& Brown,
2013). The advantage of state space models is that they are flexible,
deal with measurement inaccuracy, include information from different
sources and can be used in real time.

For secondary logs the main limitation of SSM implementations is that
they ignore movement routines. For instance, humans tend to go to work
on weekdays and sleep at night. Because SSMs are based on the Markov
property, they cannot incorporate this information. In other words, the
estimated location at a time-point \(t\) is often based only upon
measurements at timepoints \(t\) and \({t-1}\). Hierarchical structuring
and conditioning on a larger context have been suggested as ways to add
periodicity to Markovian models. These solutions are often
computationally intractable or unfeasible (Sadilek \& Krumm, 2016).
Moreover, these models often assume time and space invariance
(i.e.~location is not a direct function of time or space). These
mathematical assumptions are violated in the case of human movement
patterns. For this reason we do not consider existing SSMs to be useful
for imputing missing data in smartphone mobility logs.

@@

On the other hand, a few researchers have explicitly attempted to impute
missing data from human mobility patterns (Barnett \& Onnela, 2016;
Palmius et al., 2017 ; Wu, Fu, Wang, Xiao, \& Fu, 2017). Importantly,
none of them worked with secondary logs, but rather with custom logs
that have higher sampling frequency. Nonetheless we will detail what
they did as informative examples. Palmius et al. (2017) deal with the
measurement inaccuracy in custom logs by removing from the data set all
unique low-accuracy data points with a speed greater than 100 km/h.
Subsequently the researchers down sample the data to a sample rate of 12
per hour using a median filter. Moreover, Palmius et al. (2017) explain:

\begin{quote}
\enquote{If the standard deviation of in both latitude and longitude
within a 1 h epoch was less than 0.01 km, then all samples within the
hour were set to the mean value of the recorded data, otherwise a 5 min
median filter window was applied to the recorded latitude and longitude
in the epoch}.
\end{quote}

Missing data was imputed using the mean of measurements close in time if
the participant was recorded within 500m of either end of a missing
section and the missing section had a length of \(\leq 2h\) or
\(\leq 12h\) if this occurred after 9pm. In cases where the previous
conditions are not met no values are imputed.

Barnett and Onnela (2016) follow a different approach which is, to the
best of our knowledge, the only principled approach to dealing with
missing data in human mobility data. Barnett and Onnela (2016) work with
custom logs where location is measured for 2 minutes and subsequently
not measured for 10 minutes. Barnett and Onnela (2016) handle missing
data by first converting data to mobility traces, which are defined as a
sequence of flights and pauses. Flights are segments of linear movements
and pauses corresponding to periods of time where a person does not
move. Subsequently, the authors impute missing data by:

@@

Subsequently let us simplify by assuming that a persons location is on a
two-dimensional Euclidean plane. Let a person's true location on this
two-dimensional plane be \(G(t) = [G_x(t) G_y(t)]\) where \(G_x(t)\) and
\(G_y(t)\) denote the location of the individual at time \(t\) on the
x-axis and y-axis respectively. For simplicity, we can assume that the
x-axis is the longitude and the y-axis is the latitude. Moreover, let
\(D \in \mathbb{R}^2\) be the recorded data containing the latitude and
longitude. In addition, let \(a\) denote the estimated accuracy of the
recorded data. Furthermore, \(G(t)\), \(D\) and \(a\) are indexed by
time labeled by the countable set \(t = t_1 < ... < t_{n+1}\). If
\(D_t = \emptyset\) it is considered \emph{missing}. The measure of
accuracy \(a_t\) is given in meters such that it represents the radius
of a 67\% confidence circle. The reader should keep in mind that a high
accuracy value \(a_t\) for a measurement \(D_t\) means that the
measurement is inaccurate (this can be rather confusing).

\subsection{Personalised Map Matched
Imputation}\label{personalised-map-matched-imputation}

@@

First, following Barnett and Onnela (2016) we categorise all time-points
\(t\) as either belonging to the set \(P\) (pause) or set \(F\)
(flight). Conceptually pauses can be understood as periods of time where
an individual spends significant amount of continuous time without
moving. Flights are the times where the individual is moving. Let
\(t_a\) be a pause of length \(n\). \[t_a = t_i < ... < t_{i+n}\] Let
\(t_b\) be a pause of length \(m\) such that there is no temporal
overlap between \(t_a\) and \(t_b\) and \(t_b\) is the first pause after
the end of \(t_a\): \[t_b = t_j < ... < t_{j+m}|t_{i+n}< t_j\] Then it
follows that between the two pauses there must be a flight indexed by
\(t_x\) of length \(j-i+n\). \[t_x =  t_{i+n} < ... < t_{j} |t_x \in F\]
We define pause locations \(G(t_a), G(t_b) | t_a, t_b \in P\) as
locations where an individual spends an extended amount of time in the
same space (e.g.~school, home, work, train station, barber shop, bar,
gym). Importantly, our model assumes period and cyclic human movement
such that there are many pauses \(t_{a1},t_{a2},...,t_{an}\) such that
\(G(t_{a1}) = G(t_{a2}) = ... = G(t_{an})\). Moreover,it is possible for
two consecutive pauses \(t_a\) and \(t_b\) to share a location
\(G(t_a) = G(t_b)\) such that \(t_a \ne t_b\). For example, if the
individual leaves home for a run and returns home without stopping
anywhere else.

@@

\subsection{Personalised Map Matched Imputation
algorithm}\label{personalised-map-matched-imputation-algorithm}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \emph{Map building}: Extract from measurements \(D\) all pause
  locations and path locations to create a personalised map consisting
  of discrete bins.
\item
  \emph{Binning}: Assign each measurement \(D\) to a unique discrete
  location.
\item
  \emph{Imputing}: Use a classification method to predict missing
  measurements based on all the available information.
\end{enumerate}

@@

Following Wu et al. (2017)'s spatial-temporal-semantic (STS) feature
extraction algorithm our aim is to transform pause and path locations
into machine learning friendly discrete locations, which we will call
location bins. There are multiple ways of extracting such measurement
clusters in the literature, such as Spatio-Temporal Density-Based
Spatial Clustering of Applications with Noise (ST-DBSCAN) and sequence
oriented clustering (SOC) (Birant \& Kut, 2007). We will focus on
methods which explicitly deal with mobility patterns extracted from
smartphone logs (Barnett \& Onnela, 2016; Palmius et al., 2017 ). Both
of these methods pre-process the data and subsequently use two steps to
extract pause locations: first they extract pauses and their
corresponding locations, then they cluster pause locations based on
spatial proximity. This implementation of PMMI uses a stricter version
of Barnett and Onnela (2016)'s approach to extract pauses.

@@

All the code is available on a public repository\footnote{\href{https://github.com/sobradob/thesis}{Github}
  (Sobrado, 2018)} .
